---
title: "Emma Levin CEE599 Pset 1"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
library(vioplot) # load "vioplot" package
```


1. Write an R function that computes the third largest value in a vector of numbers. Your function should report an error if there are fewer than three numbers in the vector or if the input is something other than a vector of numbers.

```{r}
largest3 <- function(vector) {
  # print error if there are fewer than three numbers in the vector 
  if (length(vector) < 3) {
    print('Error: input vector has fewer than three elements')}
  # print error if the input is something other than a vector of numbers
  if (!is.numeric(vector)) {
    print('Error: input vector is something other than a vector of numbers')}
  
  # to calculate 3rd largest value in vector:
  # remove 1st and 2nd max values incrementally 
  first_max = max(vector)
  vector_no_1max = vector[vector != first_max]
  
  second_max = max(vector_no_1max)
  vector_no_1max_2_max = vector_no_1max[vector_no_1max != second_max]
  
  # return the third max value in of the vector
  third_max = max(vector_no_1max_2_max)
  return(third_max)
}

# test cases 
vector1 = c(1,2,3,4,5,6,7) # answer = 5
largest3(vector1)

vector2 = c(1,2) 
# should get: Error: input vector has fewer than three elements
largest3(vector2)

vector3 = c('hat', 'dog', 'cat', 'apple') 
# should get: Error: input vector is something other than a vector of numbers
largest3(vector3)

```

2. Write a function kmax(x, k=1, …) that returns the kth largest element of x. When k=1, your function should behave just like max(). Your function should report an error when the inputs provided by the user are not realistic / feasible.

```{r}
kmax <- function(vector, k) {
  # print error if there are fewer than three numbers in the vector 
  if (length(vector) < k) {
    print('Error: input vector has too few elements to compute kmax')}
  # print error if the input is something other than a vector of numbers
  if (!is.numeric(vector)) {
    print('Error: input vector is something other than a vector of numbers')}
  
  # following the methods above for 3rd max, compute 1st, 2nd, ... max and 
  # keep removing them from the vector until we are left with kth max
  for (i in 1:k){
    ith_max = max(vector)
    vector = vector[vector != ith_max]
  }
  # return kth max
  return(ith_max)
}

# test cases 
vector1 = c(1,2,3,4,5,6,7) 
kmax(vector1, 2) # answer = 6
kmax(vector1, 1) # answer = 7
kmax(vector1, 4) # answer = 4
kmax(vector1, 8) # Error: input vector has too few elements to compute kmax

vector2 = c('hat', 'dog', 'cat', 'apple') 
# should get: Error: input vector is something other than a vector of numbers
kmax(vector2, 3)
```

3. Annual maximum flood flows in the Po River at Pontelagoscuro, Italy over a 61- year period are given  (in m3/s).

```{r}
# save data for problems 3 and 4 as csv and load as a dataframe 
df <- read.csv(file = "pset1.csv")
head(df,n=3) 
```

1) compute the sample mean and standard deviation
```{r}
mean(df$Po) # mean
sd(df$Po) # sd
```

2) plot a histogram
```{r}
# number of bins should be approximately equal to sqrt(n) (number of datapoints)
nbins = round(sqrt(length(df$Po)), digits=0)
# plot histogram
hist(df$Po, breaks = nbins, col = "lightblue", border = "pink", main = "Histogram of Po River maximum flood flow", xlab = "maximum flood flows (m^3/s)")
```
3) compute the quartiles and plot a box and whiskers plot, and comment on the distribution
```{r}
# quartiles of distribution
quantile(df$Po,probs=c(0,0.25,0.5,0.75,1))

# box and whiskers plot 
p<-boxplot(df$Po ,
           xlab = 'Po River',
           ylab = "Drainage area (mi^2)", 
           main = "maximum flood flow (m^3/s)", #Add Title
           ylim=c(min(df$Po),max(df$Po)), #Set plot limits
           outline=F, #Remove Outliers
           whiskcol=NA, #Remove original 
           staplecol=NA #Remove Staples
)

#Add grid lines if desired
grid(nx = NULL, ny = NULL,
     lty = 1,      # Grid line type
     col = "gray", # Grid line color
     lwd = 1)      # Grid line width

#Next loop through the x-axis groups to calculate and plot segments for 5th and 95th quantile
w = 0.1 #staple width

q <- quantile(df$Po,p=c(0.05,0.95)) #Get 5th and 95th values
segments(x0 = 1, y0 = q[1], x1 = 1, y1 = q[2],lwd=2) #wiskers
segments(x0 = 1-w, y0 = q[1], x1 = 1+w, y1 = q[1],lwd=2) #staples for 5th percentile
segments(x0 = 1-w, y0 = q[2], x1 = 1+w, y1 = q[2],lwd=2)  #staples for 5th percentile

#Add boxplot back on top since base R plots are based on code order for plot elements
boxplot(df$Po, 
        outline=F, #Remove Outliers
        whiskcol=NA, #Remove original 
        staplecol=NA, #Remove Staples
        add=T # This is required to add on top
)
```
The distribution of maximum flood flow at the Po river appears to be fairly symmetric, since the mean and median are roughly equivalent (~5400 m^3/s). According to the histogram, the distribution is unimodal. According to the box and whisker plot, the 10%-90% quantiles are approximately 2600-8000 m^3/s, which encompasses a typical range of maximum flood flows at this river. 

Flood embankments along the banks of the river can withstand a flow of 5000 m3/s. What is the probability that this will be exceeded during a 12-month period?
```{r}
# Use the empirical cdf function P(x>5000) = 1-P(x<5000)
# where P(x<5000) is the quantile of the data at 5000 m3/s (also known as the cdf)

# create function for given data
F_cdf <- ecdf(df$Po)

# Get the quantile for flow of 5000 m3/s
quantile_5000 <- F_cdf(5000)

# flow that exceeds
1 - quantile_5000

```
The probability that a flow of 5000m3/s will be exceeded during a 12-month period is 0.59. 

4. We are given the annual maximum flows in m3/s in the Colorado River at Black Canyon for a 52-year period:

 1) compute the sample mean and standard deviation

```{r} 
# remove NA values from column with Co river data
Co <- na.omit(df$Co)

mean(Co) # mean
sd(Co) # sd
```

 2) plot a histogram 
```{r}
# number of bins should be approximately equal to sqrt(n) (number of datapoints)
nbins = round(sqrt(length(Co)), digits=0)
# plot histogram
hist(Co, breaks = nbins, col = "lightblue", border = "pink", main = "Histogram of Co River maximum flood flow", xlab = "maximum flood flows (m^3/s)")
```
 
 3) compute the quartiles and plot a box and whiskers plot, and comment on the distribution.
```{r}
# quartiles of distribution
quantile(Co,probs=c(0,0.25,0.5,0.75,1))

# box and whiskers plot 
p<-boxplot(Co ,
           xlab = 'Co River',
           ylab = "Drainage area (mi^2)", 
           main = "maximum flood flow (m^3/s)", #Add Title
           ylim=c(min(Co),max(Co)), #Set plot limits
           outline=F, #Remove Outliers
           whiskcol=NA, #Remove original 
           staplecol=NA #Remove Staples
)

#Add grid lines if desired
grid(nx = NULL, ny = NULL,
     lty = 1,      # Grid line type
     col = "gray", # Grid line color
     lwd = 1)      # Grid line width

#Next loop through the x-axis groups to calculate and plot segments for 5th and 95th quantile
w = 0.1 #staple width

q <- quantile(Co,p=c(0.05,0.95)) #Get 5th and 95th values
segments(x0 = 1, y0 = q[1], x1 = 1, y1 = q[2],lwd=2) #wiskers
segments(x0 = 1-w, y0 = q[1], x1 = 1+w, y1 = q[1],lwd=2) #staples for 5th percentile
segments(x0 = 1-w, y0 = q[2], x1 = 1+w, y1 = q[2],lwd=2)  #staples for 5th percentile

#Add boxplot back on top since base R plots are based on code order for plot elements
boxplot(Co, 
        outline=F, #Remove Outliers
        whiskcol=NA, #Remove original 
        staplecol=NA, #Remove Staples
        add=T # This is required to add on top
)
```
The distribution of maximum flood flow at the Co river appears to be positively skewwed, since the mean is greater than the median (2837.5 m^3/s > 2410.0 m^3/s). According to the histogram, the distribution is unimodal. According to the box and whisker plot, the 10%-90% quantiles are approximately 1700-5500 m^3/s, which encompasses a typical range of maximum flood flows at this river. 

How does this distribution different from that of Part 3?
```{r}
# plot the violin plots next to each other on same axis for better comparison
vioplot(df$Po, Co, names = c("Po River", "Colorado River"), col = "lightblue")
# Add title and axis labels
title(main = "Maximum River Flows", xlab = "River", ylab = "Max flow (m3/s)")

#Add grid lines 
grid(nx = NULL, ny = NULL,
     lty = 1,      # Grid line type
     col = "gray", # Grid line color
     lwd = 1)      # Grid line width

```

This Colorado River distribution is less symmetric than that of part 3 (positively skewed, while the distribution of part 3 is symmetric). This distribution is also centered at lower values than that of distribution 3 (mean= 2837.5 m^3/s vs. 5408.197 m^3/s and median = 2410.0 vs. 5400 m^3/s), which means that the Po river typically experiences a higher annual maximum flood flow. Lastly, given the outlier of the distribution for the Colorado river (8500 m^3/s), this distribution has a higher range of annual maximum flood flows than that of the Co river. However, the standard deviation, which is less sensitive to outliers, is higher for the annual maximum flood flows for the Colorado river, meaning the data are more concentrated around their center. 

